---
title: "Using spatially explicit models to improve sampling of rare marine species: species-specific behavior and ecology drive depth distribution of cetacean eDNA in pelagic marine environments"
author: "Amy M. Van Cise^1\\*^, Eiren K. Jacobson^2^, Megan R. Shaffer^3^, Pedro F.P. Brandão-Dias^3^, David L. Miller^4,5^, Brice X. Semmens^6^, Len Thomas^2^, Ryan P. Kelly^3^, A. Ole Shelton^7^, Krista M. Nichols^7^,  Kim M. Parsons^7^"
format: html
editor: visual
execute:
  echo: false
  message: false
  warning: false
project:
  output-dir: docs
bibliography: references.bib
csl: molecular-ecology.csl
---

1.  School of Aquatic and Fishery Sciences, University of Washington, Seattle, WA, USA
2.  Centre for Research into Ecological and Environmental Modelling, School of Mathematics and Statistics, University of St Andrews, St Andrews, Scotland
3.  School of Marine and Environmental Affairs, University of Washington, Seattle, WA, USA
4.  Biomathematics and Statistics Scotland, Dundee, UK
5.  UK Centre for Ecology and Hydrology, Lancaster Environment Centre, Lancaster, UK
6.  Scripps Institution of Oceanography, University of California San Diego, San Diego, CA, USA
7.  Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA

<br>

^\*^Corresponding author email: avancise\@gmail.com

<br>

Running page head: Depth distribution of rare target detections

```{r}
library(tidyverse)
library(patchwork)
library(cowplot)
library(pander)
library(PNWColors)
library(kableExtra)

# load data
load("ProcessedData/detect_data_clean.RData")
load("ProcessedData/potential_whalefall.RData")
load("ProcessedData/H1models.RData")
load("ProcessedData/m1.2clean_performance.Rdata")
load("ProcessedData/m1.2performance.Rdata")
load("ProcessedData/m3.0models_preds_0.05degree.Rdata")
load("ProcessedData/H3.0c_pred_flt_clean.Rdata")
load("ProcessedData/m3.0cperformance.Rdata")
load("ProcessedData/POD_est_per_samp.Rdata")

# load plots
load("Figures/sampling_map.Rdata")
load("Figures/detectDepth_plots.Rdata")
load("Figures/H1PODplots.Rdata")
load("Figures/H3.0c_map_clean.Rdata")
load("Figures/Q2_DetectabilityByPrimer.RData")
load("Figures/Q2_TechRepsXPrimer.RData")
load("Figures/Q2_detect_probability_plots.Rdata")

metadata <- read.csv("./Data/Hake_2019_metadata.csv")
```

## Abstract

Detecting rare species is a long-standing challenge in ecological studies that is exacerbated for many marine species due to their remote habitats and elusive behaviors. Environmental DNA (eDNA) sampling is a powerful standalone or complementary tool that promises to improve spatiotemporal models of species' distribution and abundance, but positive detections of rare species remain difficult to obtain. Targeting optimal sampling depths can improve the detection rates of cetaceans, which are highly mobile and have species-specific depth ranges spanning the surface to 3 km. In this study, we examine the depth distribution of eDNA detections from 16 cetacean species in the California Current Ecosystem using systematic eDNA samples collected from the surface to 500 m. We sequenced five replicates each of 516 samples from 175 unique geographic locations and up to 5 depth strata, obtaining 468 positive cetacean detections across 2,580 sequencing replicates. Given the presence of cetacean eDNA in a sample, we found that the probability of detecting that eDNA using genetic metabarcoding was driven by primer specificity, technical replication, and the number of freeze/thaw cycles a sample was subjected to. Futher, species-specific eDNA detection rate varied with depth and generally reflected known species-specific use of the water column, but could also be modeled using information on taxonomy or diet. Most dolphin and porpoise species had highest predicted detection rates at the surface, whereas most baleen whale species had highest detection rates at the surface in regions close to shore and at 500 m off the continental shelf. Beaked whales had highest detection rates between 300 and 500 m, and were unlikely to be detected on the continental shelf. The depth maximizing detection rate for each species was not geographically consistent, indicating species-specific geographic variability in depth distribution. Our results indicate that the challenge of detecting cetacean species using eDNA can be mitigated in the experimental design phase by using niche-based or spatial distribution models to inform sampling strategies, especially sampling depths. eDNA sampling strategies for cetaceans should account for species-specific depth distributions and habitat use to optimize detection rates; incorporating depth-informed sampling may increase detection rates over non-depth-targeted surveys and strengthen our ability to estimate marine mammal distribution and abundance.

## Introduction

Detecting rare species is a longstanding challenge in many ecological fields, e.g., ecosystem-level studies of biodiversity or community composition [e.g., @cao_etal01; @cao_etal01; @queheillalt_etal02; @white_etal23; @jeliazkov_etal22] as well as species-level studies of spatial ecology [e.g., @engler_etal04; @lelay_etal10; @guisan_etal06]. Many methodological approaches have been developed to address this problem, both in the experimental design phase [@lelay_etal10; @guisan_etal06] and in the data analysis phase [@engler_etal04; @guelat_kery18; @lyet_etal13]. Recently, environmental DNA (eDNA) has emerged as a viable tool for detecting rare species, with promise to improve spatial distribution models (SDMs) and biodiversity monitoring [e.g., @afonso_etal24; @valdivia-carrillo_etal25], particularly when integrated with existing survey methods [e.g., @claver_etal25; @shelton_etal22].

While eDNA-based species detection is a promising approach to estimating or predicting the abundance and distribution of rare species, it may be limited by the collection of point samples of relatively small amounts of water (generally 1-10 L) throughout the study area. This necessitates careful consideration of the inferential aims of the study when deciding where to collect samples to determine species presence/absence or describe geographic ($x$ and $y$) distribution, either independently of or alongside depth ($z$) distribution. On the other hand, explicitly sampling across the $z$ dimension, which is often ignored in marine mammal monitoring programs, will enhance our capability to describe spatiotemporal variability in $z$ habitat use that, when combined with variability in $xy$ habitat use, may provide novel insight into the spatiotemporal ecology of these species in three dimensions to inform both the management of rare species and future sampling designs based on 3D ecological behaviors.

Marine mammals, and specifically cetaceans, are notably some of the rarest species in the marine environment in terms of in terms of their relative abundance and the patchiness of their distribution, with estimates of cetacean density [e.g., @Becker2020; @paiu_etal24] several orders of magnitude smaller than that of fishes, plankton, or microbes [e.g., @bar-on_etal18; @irigoien_etal14; @whitman_etal98] and decreasing with individual biomass as predicted by ecological theory [@jennings_etal08]. Traditionally, cetacean habitat use and distributions are mapped using observations from visual and/or passive acoustic surveys [e.g., @dalpaz_etal21; @fiedler_etal23; @Forney1998; @Findlay1992; @becker_etal22; @kanaji_miyashita21]. Visual line transect surveys are the most common, in which observers on board a ship or airplane record detected marine mammals and their distance to the transect line, covering the survey area in a systematic pattern. These surveys are limited to surface (typically \<10 m) observations, where most cetaceans spend very little of their time, made in a relatively small radius from the observation platform [e.g., approximately 11 km from a typical NOAA survey vessel; @Barlow2016], and during daylight hours - and their effectiveness decreases in poor weather or sea state. Ship-based passive acoustic line-transect surveys tow hydrophones along similar systematic transects, and can detect mid-to-high frequency signals within a radius of 10 km, and lower frequency signals in a radius of 10s-100s of km. Static passive acoustic data collection employs hydrophones mounted to surface buoys or the seafloor, with similar detection frequencies and sensitivities. Passive acoustic surveys are not limited to surface or daylight observations, but only detect species that are actively vocalizing.

Systematic eDNA surveys, where water samples are collected along transects, do not share limitations related to depth, daylight, weather conditions, or species' vocal activity and are therefore potentially powerful as standalone or complementary monitoring tools [@pikitch18]. However, it is important to consider the unique limitations of eDNA surveys when using this approach independently or jointly with other methods. Because eDNA survives for several hours to days [@brandao-dias_etal25], an eDNA sample represents a longer temporal integration of species’ occurrence at a given location than visual or acoustic snapshots. Additionally, the distribution of eDNA is influenced by diffusion and transport associated with oceanographic processes [@yamahara_etal25a; @brasseale_etal25; @xiong_etal25], as well as sinking rates, and therefore eDNA detected in a sample may have a provenance up to 10s of km from the sample site [@Andruszkiewicz2019]. Despite this temporal persistence and potential for movement, eDNA has been previously shown to reflect geographic distributions and depth patterns of marine species [@canals_etal21; @guri_etal25; @shelton_etal22].

Marine surveys are expensive and logistically challenging, and eDNA samples represent a small volume of water (typically \<3 L per sample) compared to the survey area. They are also costly to process, so studies targeting a specific set of rare species may benefit from restricting their range of sampling depths via spatial or niche-based modeling. There is currently a lack of published data regarding optimal sampling depths to target rare taxa, or what ecological and observational processes may drive variability in the probability of detecting those taxa across depths.

Here, we first quantify the probability of detecting cetacean eDNA present in a sample, then describe the $z$ distribution of cetacean species' eDNA in the water column and use the results to identify broad ecological drivers of observed patterns. We specifically ask:

1.  Does probability of detection (POD) of cetacean eDNA vary in response to technical replication [@ficetola_etal15], primer specificity [@shaffer_etal25], or post-extraction DNA degradation [@brandao-dias_etal25] due to laboratory freeze/thaw cycles?
2.  Does detection rate of cetacean eDNA vary in the depth ($z$) dimension? If so, can this variability be predicted by preferred prey categories or phylogeny (e.g. taxonomic family)?
3.  Does latitudinal and longitudinal variability in eDNA distribution interact with depth variability to affect species-specific detection rate in three dimensions?

To answer these questions, we leverage an opportunistic set of eDNA samples collected from seawater in the California Current in 2019 in order to assess the abundance of North Pacific hake (*Merluccius productus*). We examine cetacean detections generated via genetic metabarcoding of three mitochondrial target regions. We first use an occupancy model implementend in a hierarchical Bayesian framework to examine the POD of cetacean DNA given it's presence in an eDNA sample. Next, we use a GAM framework to examine species-specific detection rate across sampling depth (0-500 m), then incorporate spatial location information to examine species-specific variability in detection rate in three dimensions.

## Methods

{{< include zDist_methods.qmd >}}

## Results

{{< include zDist_results.qmd >}}

## Discussion

Cetaceans are highly mobile and rare among marine species in abundance and density. Moreover, the distribution of cetaceans throughout the water column (their z distributions), spanning 100s to 1000s of meteres, is challenging to characterize, often limited by traditional survey approaches. eDNA sampling at depth addresses this data gap. Mapping the $z$ distribution of cetacean eDNA has the potential to provide new insight regarding their 3D spatial use and improve estimates of abundance or distribution, especially for visually or acoustically elusive species. Here we show that the use of occupancy models and niche-based GAMs can inform and optimize eDNA sample collection and processing. For cetaceans and other highly mobile rare marine species, tailoring experimental design based on the results of this study could increase detection rates and improve distributional or abundance models while minimizing cost.

#### Question 1:

We began with an occupancy model to explicitly separate the effects of cetacean eDNA presence at the sampling site/in the bottle and downstream detection in the sequence data. By implementing this framework we observed an increase in POD with increased primer specificity. Similarly, increasing technical replication increases POD, except that increasing freeze/thaw cycles decreases POD and may counteract the effect of technical replication for longer target regions if technical replicates are not processed concurrently, as was observed with DL in this study. Shorter regions (MFU and MV1) are more robust to decay from freeze/thaw cycles but lower genetic variability in cetacean DNA in each of those regions hinders the ability to classify sequences to species, which necessitated manual curation of the reference database for each of these primers. Within the Delphinidae family, several sequences could not be resolved to species, resulting in the loss of 2 MFU species-level detections and 17 MV1 species-level detections. These family-level detections could not be included in downstream species-specific models of eDNA distribution.

Similar studies have found that observation bias and rare species dropout in eDNA sampling are affected by a number of parameters in the sample collection and processing protocols [e.g. @shelton_etal23; @gold_etal23], including sampling volume (Patin et al. *in review*), filter type and pore size [@liu_etal24], primer specificity [@shaffer_etal25], and technical replication [@ficetola_etal15], among others. The results of our study indicate that when present in a sample, rare species can be detected more efficiently using primers with higher specificity and targeting regions with greater variability among the taxa of interest; further, the samples should be subjected to a limited number of freeze/thaw cycles in order to preserve the integrity of longer DNA fragments. Because increased technical replication is advisable to increase detection of rare species DNA in a sample, the best practice would be to conduct all technical replicates for a given sample set concurrently, thus only needing to thaw the extracted DNA 0-1 times before amplification and sequencing.

After accounting for observation bias in our detection rates, we found that predicted cetacean eDNA distribution across depth was qualitatively similar when fitted in an occupancy model context (Eq 1-3) as compared to the non-species specific GAM (Eq 4) in Question 2. Estimates of cetacean POD from the occupancy model were higher overall than detection rate estimates from GAM models, with confidence intervals approaching 1. This is logical given that the Bayesian hierarchical occupancy model estimates POD conditional on the presence of target eDNA in the sample, whereas the GAM did not distinguish POD from the probability of a target species' occurrence at a sampling site, leading to lower overall detection rate. Continued eDNA sampling at these locations will facilitate the convergence of a fully hierarchical occupancy model with species-specific splines on depth; improving our ability to delineate the effects of observation error (eDNA POD) from ecological processes (eDNA occurrence) in detection rate.

#### Question 2:

Detection rate across all cetaceans varied with sampling depth (Supplemental Figure S1) and was best described at a species-specific level, broadly reflecting our current understanding of species-specific variability in use of the water column (@fig-H1POD-species). This was especially evident among the delphinids and porpoises, with some noticeable deviations from expected trends in both the baleen whales and the beaked whales. These trends could be inherent to the behavior of the animals themselves, or more likely may reflect the "behavior" of eDNA.

For example, baleen whales are generally thought to be shallow divers, and while some dives up to approximately 500 m have been recorded in a few species [e.g., fin whales, @panigada_etal99], recent reports of time-at-depth for cetacean species indicates that baleen whales spend the vast majority of their time within the top 100-200 m of the ocean [@watwood_buonantony12; @u_s_departmentofthenavy17]. We observed a sharp spike in detection rate at 500 m for two of the three baleen whale species included in our model predictions. Gray whale eDNA was not likely to be observed at 500 m; this is most likely because they are a nearshore coastal species that is rarely observed over deep water. Humpback and fin whales both have pelagic distributions, and both species' eDNA was as likely to be observed at 500 m as at the surface, with low detection rates at intermediate depths.

We considered several possible explanations for this observed pattern. We first considered the possibility that some of these deeper eDNA detections, especially those of deeper-diving fin whales, are detections of deep dives made by animals foraging for krill at depth. However, based on our current knowledge of the diving limits of baleen whales, we consider it highly unlikely that most of the deep detections are indicative of the presence of the detected species at that depth. We then considered the possibility that the large spike in deep detections was driven by whale fall or predation (i.e. whole or parts of animals sinking in the water column). In total, baleen whales were detected below 250 m at 28 locations; of those, 8 detections (28.5%) were within 100 m of the ocean floor and might be considered potential detections of whalefall. A map of these detection locations is included in Supplemental Figure S4. In addition to this, a single baleen whale detection coincided with the detection of killer whale eDNA belonging to a potential baleen whale predator. No eDNA detections of baleen whales coincided with the detection of shark eDNA in the data generated from MFU or MV1 sequencing runs.

Finally, we considered the possibility that these deep-water detections are attributable to sinking whale feces. Even after accounting for whale fall or predation events as possible explanations, there remain 19 deep-water detections of baleen whale eDNA (68% of deep-water detections) that are unexplained. While sinking rates and depths of baleen whale feces are generally unknown, we know from expert observations that fecal material from whales is quite large and can sink rapidly from the surface (K.M. Parsons, unpublished observations). Research on oceanic particle sinking rate indicates that smaller particles (up to 6 mm) can sink at rates of \~50-300 m day^-1^, and that sinking rates increase rapidly with particle size while degradation rates decrease with increasing particle size, with maximum sinking depths in the range of \~800-5,000 m day^-1^ [@giering_etal20; @mcdonnell_etal15; @iversen_ploug10]. eDNA can be detectable for up to 2-4 days at the surface [@brandao-dias_etal25; @wood_etal20], and likely longer at depth where it is colder, darker, and there are fewer microbes digesting cellular material [@andruszkiewicz_etal17].

Based on this, the maximum sinking depth of larger poop particles may far exceed the $z$ range of our study area. It is possible that baleen whale detection rates will continue to increase with increasing depth in some regions due to rapid particle sinking and longer eDNA residency times at depth [@andruszkiewicz_etal17]. Generally, eDNA is thought to travel 2-20 horizontal km before it is no longer detectable [@Andruszkiewicz2019; @baker_etal18; @silva_etal25]; therefore on a broad scale these detections are likely to reflect the $xy$ distribution of these species but the temporal resolution of detections may decrease from days to weeks in samples from deeper environments.

Beaked whales were also generally detected deeper than where time-at-depth models predict they spend most of their time (@fig-H1POD-species), although still within their range of vertical habitat use. Beaked whales spend less of their time near the surface than either baleen whales or delphinids; time-at-depth models estimate that they spend approximately 70% of their time below 100 m [@watwood_buonantony12; @u_s_departmentofthenavy17], with depth ranges spanning down as far as 3,000 m [@schorr_etal14]. While below the surface, beaked whales generally make deep (500+ m) foraging dives interspersed with several shallow (200-300 m) "resting" dives [@baird_etal06]. We found that goose-beaked whales, considered the most common beaked whale in the Pacific Ocean, were most likely to be detected between 200 and 300 m, in the range of their resting dives. Northern giant bottlenose whales were most likely to be detected at 400-500 m; as 500 m was the extent of our sampling range, it is possible that this species or other beaked whales have higher detection probabilities at depths greater than 500 m. While we did not make predictions of detection rate by depth for the saber-toothed whale, the three detections of this species were at 300 m (n = 1) and 500 m (n =2).

Finally, for both baleen whales and beaked whales, high detection rates at depth may be caused by either slower decay and therefore longer residency times or decreased abundance of highly prevalent taxa such as microbes, plankton, fishes, etc [@robinson_etal24; @pinfield_etal19] that can "swamp out" the signal from rarer taxa at shallower depths. This latter issue is likely to be exacerbated when using general primers; in our study, we expect that our cetacean detections from MV1 and MFU primers were more susceptable to a swamping effect than DL, which is designed to exclude non-cetacean DNA and preferentially amplify cetaceans.

In general, sampling near the surface will target delphinids and porpoises, and sampling at 300+ m will target beaked whales, with baleen whales detectable at both the surface and depths \>300 m. While we found that detection rate varies by species, we also observed significant trends in detection with depth when species were grouped by phylogenetic family or by broad prey category. This indicates some degree of similarity in habitat use across the $z$ axis within these groups, such that researchers wanting to design an eDNA monitoring study targeting a species for which there is little information could stratify their sampling according to phylogenic or niche-based distribution models.

#### Question 3:

Species-specific maximum detection rate depths varied geographically, as illustrated by the three example species used in this study (@fig-H3PODxy). At the broad scale, we expect that this pattern partially reflects geographic variability in species-specific use of the water column. That may reflect geographic shifts in the depth distribution of prey or preferred prey items of a given predator, which may in turn reflect geographic variability in oceanographic conditions.

It is also likely that some of the observed variability is directly linked to oceanographic conditions, and its effect on the fate and transport of eDNA once deposited in the water column. eDNA is generally considered observable at the surface for 2-4 days [@brandao-dias_etal25; @wood_etal20], and degradation and residence time will be affected by water temperature, pH, and microbial density [@lamb_etal22; @jo26]. During that time, horizontal transport will be affected by currents and other mesoscale processes, ane vertical transport will be affected by particle size, stratification/mixed layer depth, and upwelling, among others [@Andruszkiewicz2019]. This study is relatively insensitive to horizontal transport of eDNA, which is expected to travel 2-20 km before it is no longer detectable [e.g., @xiong_etal25; @brasseale_etal25; @brandao-dias_etal25; @silva_etal25]. Our samples were collected along transects set at 10 km intervals throughout the study area. Given this spatial resolution, our predictions of geographic variability in detection rate for rare species eDNA is most likely to be affected by eDNA sinking rates or residence time than horizontal transport.

It is likely that variability in the $z$ distribution of species-specific detection rate represents a combination of both $xy$ variability in species-specific use of the water column, as well as $xy$ variability in eDNA sinking and residency times. In both cases, underlying oceanographic processes are an important driver of the observed patterns. To assess these links between oceanography and 3D spatial distribution, our future work will explicitly focus on the development of 3D spatial distribution models using eDNA and using these models to determine the important oceanographic drivers of cetacean eDNA in 3D space.

#### Considerations for beaked whales

Among cetaceans, beaked whales were the most rarely observed in our dataset, echoing the results of previous visual and acoustic data collection efforts in the California Current that have resulted in poor performance metrics for SDMs and density estimates for this clade [e.g. @Becker2020]. For example, during regular visual surveys of the U.S. EEZ portion of the California Current ecosystem, northern giant bottlenose whales were observed approximately 45 times between 1991 and 2018 across a study area roughly double the latitudinal width of our study area [@carretta_etal23].

We observed the same species 14 times in a single survey using eDNA, and note that all but one of these detections were made below the surface (Figure 1). Most detections were made below 100 m, and predicted detection rate for this species was highest at 300+ m both with and without $xy$ model covariates. Across all beaked whales, the highest predicted detection rates occurred at depths \> 200 m (Figure 4). Incorporating $xy$ covariates illustrated that beaked whales are not likely to be observed on the continental shelf, supporting previous SDMs showing their absence near the coast [@Becker2020], and that their offshore eDNA detections are most likely to be at depths \> 300 m throughout the California Current. In the context of beaked whale ecology, this result is not surprising, but it illustrates the promise of eDNA to supplement our current database of positive detections in the case of extremely rarely sighted species such as beaked whales, which may improve SDMs considerably. eDNA can also be used to refine our understanding of habitat use in the $z$ dimension for these deep water species.

#### Marine Mammal monitoring: lessons for eDNA experimental design

Species-specific spatial models, niche-based modeling, or phylogeny-based distribution models can be used to stratify eDNA sampling designs in order to increase sampling efficiency and positive detections of target taxa, especially for rare and mobile marine species, thus improving our capacity to generate accurate estimates of cetacean abundance and distribution for use in management decisions. While eDNA sampling is promising as a complementary or standalone approach to characterizing the distributions of marine mammals in general, its efficacy will be greatest for species with SDMs that are low resolution or have large confidence intervals due to a paucity of visual or acoustic observations. These are generally deeper-diving species such as beaked whales. Future marine mammal surveys may consider integrating collection of eDNA samples from depths \>300 m to increase positive detections of these species. Increasing positive detections for these or any rare species will increase spatiotemporal resolution of SDMs and habitat use models, which can then be fed back into experimental design for future eDNA sampling surveys.

Once samples are collected, primer specificity, DNA degradation associated with freeze–thaw cycles, and the level of technical replication all influence eDNA POD in each sample when using a metabarcoding approach. Increasing primer specificity and technical replication increases POD of rare species in a sample. While targeting shorter fragments that are robust to DNA decay from freeze/thaw cycles may be a useful way to increase technical replication, it may also decrease the ability to classify sequence data down to species-level resolution. Therefore, sequencing longer and more specific fragments first, using as few freeze/thaw cycles as possible, and conducting all technical replicates simultaneously will increase the POD of DNA that is rare in the sample.

Integrating optimized eDNA sampling design with strategies that maximize DNA capture has the potential to substantially expand the utility of eDNA for modeling the abundance and distribution of rare marine species, including cetaceans. The framework developed here provides practical guidance for tailoring eDNA sample collection and processing to single or multiple taxa in accordance with study objectives, increasing detection efficiency while reducing sampling effort and cost.

## Data availability

All raw sequence data have been uploaded to GenBank (Accessions: XXXX). Detection data, sample metadata, and all code used to generate this manuscript are available at https://github.com/MMARINeDNA/zDistribution.

## Acknowledgements

The authors gratefully acknowledge the survey expertise and support of NOAA's Pacific hake survey team and the personnel of the NOAA Ship Bell M. Shimada for supporting the collection of eDNA samples on the US West Coast Hake Survey 2019. Thanks to Ana Ramon-Laca and Abi Wells in their coordination for the collection and extraction of environmental DNA samples during this survey. This material is based upon research supported by the Office of Naval Research under Award Number (N00014-22-1-2719). The authors acknowledge Research Computing at the James Hutton Institute for providing computational resources and technical support for the “UK’s Crop Diversity Bioinformatics HPC” (BBSRC grants BB/S019669/1 and BB/X019683/1), use of which has contributed to the results reported within this paper [@percival-alwyn_etal25a]. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

## Supplemental Materials

Table S1. Sample metadata.

Table S2. Mathematical notation for all variables included in the occupancy model and GAM framework in this study.

```{r}
#| label: tbl-m12AUC
#| eval: FALSE
#| tbl-cap: "Table S3. Per-species AUC, optimized TSS, and optimized TSS threshold for Q1 and Q3 models."

# m3.0c_metrics <- AUC_df_m3.0c %>% 
#   summarize(AUC = mean(AUC)) %>% 
#   bind_cols(TSS_df_m3.0c %>% arrange(desc(mean)) %>% slice_head()) %>% 
#   select(AUC, threshold, mean) %>% 
#   rename("TSS" = mean, "Threshold" = threshold) %>% 
#   mutate(species_model = "Q3")
# 
# model_metrics <- species_metrics %>% 
#   mutate(species_model = paste0(species, "-Q1")) %>% 
#   select(-species) %>% 
#   rename("AUC" = meanAUC, "Threshold" = meanThreshold, "TSS" = meanTSS) %>% 
#   bind_rows(m3.0c_metrics)
# 
# pander(m3.0c_metrics)
```

```{r}
#| label: fig-S1-null
#| fig-cap: "Supplemental Figure S1. Q1 null hypothesis plot."

m1.0POD

```

```{r}
#| label: fig-S2-TSS
#| fig-cap: "Supplemental Figure S2. TSS plots for the best-fit Q@ model (top) and Q3 model (bottom) showing the 10-fold mean TSS values across all detection thresholds. The optimal detection threshold is that which maximizes TSS."

TSS_plot / TSS_plot_m3.0c

```

```{r}
#| label: fig-SuppH1POD_obs
#| fig-cap: "Supplemental Figure S3. Q1 best fit model with observed detections."
#| fig-width: 15
#| fig-height: 7

m1.2PODwObs

```

```{r}
#| label: fig-S3-whalefall
#| fig-cap: "Supplemental Figure S4. Locations of eDNA detections within 100 m of the ocean floor, which may represent nearby whalefall."

whalefall_map
```

```{r}
#| label: fig-SuppH3PODxy_baleen
#| eval: FALSE
#| fig-cap: "Supplemental Figure S5. Depth of maximum probability of detection (POD; top) and range of depths falling within 50% CI of the depth of maximum detection (bottom) for humpback whales, separating likely whale poop (detections > 200 m) from detections within the normal depth range of humpback whales."
#| fig-width: 7
#| fig-height: 15

#depthPODmax_baleen

```
