---
title: "Species-specific behavior and ecology drive depth distribution of rare species eDNA in pelagic marine environments"
author: "Amy M. Van Cise^1\\*^, Eiren K. Jacobson^2^, Megan R. Shaffer^3^, Pedro F.P. Brandão Dias^3^, David L. Miller^4,5^, Brice X. Semmens^6^, Len Thomas^2^, Ryan P. Kelly^3^, A. Ole Shelton^7^, Krista M. Nichols^7^,  Kim M. Parsons^7^"
format: html
editor: visual
execute:
  echo: false
  message: false
  warning: false
project:
  output-dir: docs
bibliography: references.bib
---

1.  School of Aquatic and Fishery Sciences, University of Washington, Seattle, WA, USA
2.  Centre for Research into Ecological and Environmental Modelling, School of Mathematics and Statistics, University of St Andrews, St Andrews, Scotland
3.  School of Environmental and Marine Affairs, University of Washington, Seattle, WA, USA
4.  Biomathematics and Statistics Scotland, Dundee, UK
5.  UK Centre for Ecology and Hydrology, Lancaster Environment Centre, Lancaster, UK
6.  Scripps Institution of Oceanography, University of California San Diego, San Diego, CA, USA
7.  Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA

<br>

^\*^Corresponding author email: avancise\@gmail.com

<br>

Running page head: Depth distribution of rare target detections

```{r}
library(tidyverse)
library(patchwork)
library(cowplot)
library(pander)
library(PNWColors)

# load data
load("ProcessedData/detect_data_clean.RData")
load("ProcessedData/potential_whalefall.RData")
load("ProcessedData/H1models.RData")
load("ProcessedData/m1.2clean_performance.Rdata")
load("ProcessedData/m1.2performance.Rdata")
load("ProcessedData/m3.0models_preds_0.05degree.Rdata")
load("ProcessedData/H3.0c_pred_flt_clean.Rdata")
load("ProcessedData/m3.0cperformance.Rdata")

# load plots
load("Figures/sampling_map.Rdata")
load("Figures/detectDepth_plots.Rdata")
load("Figures/H1PODplots.Rdata")
load("Figures/H3.0c_map_clean.Rdata")
load("Figures/Q2_DetectabilityByPrimer.RData")
load("Figures/Q2_TechRepsXPrimer.RData")

metadata <- read.csv("./Data/Hake_2019_metadata.csv")
```

## Introduction

All analyses of community composition face a similar challenge: rare species dropout. This leads to inaccurate estimates of biodiversity or community composition and greater difficulty assessing changes in ecosystem resilience, as well as inaccurate characterization of a rare species distribution and greater difficulty assessing changes in distribution in response to environmental perturbations.

Many methodological approaches have been developed to address this problem, both in the experimental design phase (REFS) and in the data analysis phase (REFS). Statistical approaches aim to quantify and minimize the effect of rare species dropout, while sampling approaches aim to reduce the rate of dropout via robust sampling design. Within the latter category, it has been shown that adopting an ensemble approach, i.e. integrating across multiple types of observation methods in order to smooth across the weaknesses of each approach individually, often produces more robust estimates of species' distributions and community composition than using single methods alone (REFS).

eDNA is emerging as a viable tool for improving spatial distribution models for rare marine species, especially when used jointly with existing sampling methods, e.g. net tows, active acoustics, and/or bottom trawls for fishes and invertebrates. To date, these methods have been most frequently used to assess biodiversity and species' xy distributions in lakes, streams. Their use is growing in broad scale marine studies (e.g. Shelton 2019? others?), studies of 3D distributions (REF?), and tracking rare species (REF?). *I don't love this paragraph how can we make it better?*

In the case of marine mammals - notably some of the rarest species in the marine environment - two survey techniques are traditionally used. Visual line-transect surveys are the most common, in which observers aboard a ship or airplane will record detections of any marine mammals and their distance to the transect line, covering the survey area in a systematic pattern. These surveys are limited to the surface of the ocean, where most marine mammals spend very little of their time, and to daylight hours - and their effectiveness decreases in poor weather or sea state. Ship-board passive acoustic line-transect surveys collect recordings along similar systematic transects, and can often detect species up to XX kilometers away from the ship in any direction. They have some of the same limitations at visual surveys, but only detect species that are actively vocalizing. Systematic eDNA surveys, where water samples are collected along transects, do not share limitations related to depth, daylight, weather conditions, or species' vocal activity, but notably have a detection range that is much smaller than either of the two previous surveys and strongly affected by diffusion and transport by oceanographic features (REFS).

In open ocean, deep water environments, our understanding of species' distributions may be affected by sampling depth, a fundamental aspect of experimental design often overlooked in 3D environments. This is especially true for mobile species capable of utilizing a broad swath of the water column. In the pelagic marine environment, some rare species have depth distributions spanning 3km or greater (CITE Greg paper), which complicates decision-making regarding where to collect point-samples with the aim of describing x-y distribution, either independently of or alongside z-distribution.

Because eDNA samples have a limited radius of detection, and because they are costly both to collect and to process, studies targeting a specific set of rare species may benefit from restricting their range of sampling depths, especially when describing x-y distribution has a higher priority than describing z-distribution. However, there is currently a lack of published data providing guidance regarding the range of sampling depths that can be used to target certain species, taxonomic groups, or ecological groups of rare species, or what ecological or observational processes may drive variability in the probability of detecting specific species across depths.

-   \[3D novelty point\] Further, eDNA promises new capability to monitor spatiotemporal variability in vertical habitat use that, when combined with variability in x-y habitat use, may improve novel insight into habitat use for species with 3D distributions. If the vertical distribution of eDNA faithfully reflects the vertical distribution of rare species, this tool may become a viable non-invasive option for understanding 3D habitat use of rare and elusive species on a broad scale.

Here, we aim improve our understanding of the depth distribution of rare species' eDNA in the water column and identify major environmental and ecological drivers of observed patterns. We specifically ask three questions:

1.  Does cetacean eDNA distribution vary in the z-dimension? If so, what major ecological factors (e.g. preferred prey, time spent at depth, or phylogeny) can be used to predict cetacean eDNA distribution across depths?
2.  Do common sources of observational bias, e.g. technical replication, choice of primers, or sampling volume, interact with sampling depth to affect probability of detection?
3.  Does x-y variability in eDNA distribution interact with depth variability to affect probability of detection?

To answer these questions, we utilize eDNA (water) samples collected in the California Current in 2019 in order to assess the abundance of North Pacific hake\* (*Merluccius productus*). We leverage this opportunistic dataset to examine marine mammal detections generated via genetic metabarcoding of three mitochondrial target regions. By collapsing eDNA detections from these three primers across x and y dimensions and focusing only on the z-distribution of sampling effort (0-500 m), we aim to determine (1) whether species- or taxon-specific detection probability varies with depth and (2) how depth-dependent detection probability varies with primer specificity, technical replication, or DNA degradation caused by freeze/thaw cycles. Finally, we explicitly re-introduce x-y dimensions in order to determine (3) whether depth-dependent detection probability varies with latitude or longitude.

## Methods

{{< include zDist_methods.qmd >}}

## Results

{{< include zDist_results.qmd >}}

## Discussion

#### Question 1:

In addition to being extremely rare targets, marine mammals are highly mobile in xy space, and their vertical ranges span 100s to 1,000s of meters. Mapping the vertical distribution of their eDNA can provide new insight regarding their 3D spatial use, including seasonal or geographic variability in depth distribution. Additionally, eDNA monitoring has the potential to improve our understanding of xy spatial distribution, especially for visually or acoustically elusive species, when used in joint species distribution models. In these cases, optimizing the sampling depth to target the species of interest could increase detection rates and improve xy distributional models while minimizing cost.

POD varied significantly with sampling depth (Supplemental Figure S1), and was best described at a species-specific level, and broadly reflected our current understanding of species-specific variability in use of the water column (@fig-H1POD-species). This was especially true among the delphinids and porpoises; on the other hand, we observed some noticeable deviations from expected trends in both the baleen whales and the beaked whales. These trends could be inherent to the behavior of the animals themselves, or more likely may reflect the "behavior" of eDNA.

For example, baleen whales are generally thought to be shallow divers, and while some dives up to approximately 500 m have been recorded in a few species [e.g. fin whales, @panigada_etal99], recent reports of time-at-depth for cetacean species indicates that baleen whales spend the vast majority of their time within the top 100-200 m of the ocean. However, we observe a sharp spike in detection probability at 500 m for two of the three baleen whale species included in our model predictions. Gray whales were not likely to be observed at 500 m; this is most likely because they are a nearshore coastal species that is rarely observed in deep water. Humpback and fin whales both have pelagic distributions, and were both as likely to be observed at 500 m as they were at the surface, with low POD at intermediate depths.

This observed pattern has several possible explanations. It is possible that some of these deeper eDNA detections, especially those of deeper-diving fin whales, are detections of deep dives made by animals foraging for krill at depth. However, based on our current knowledge of the diving limits of baleen whales, we consider it highly unlikely that most of the deep detections are indicative of the presence of the detected species at that depth. We considered the possibility that the large spike in deep detections was driven by whalefall or predation (i.e. whole or parts of animals sinking in the water column). In total, baleen whales were detected below 250 m at 28 locations; of those, 8 detections (28.5%) were within 100 m of the ocean floor and could be potential detections of whalefall. A map of potential whalefall locations is included in Supplemental Figure S2. In addition to this, a single baleen whale detection coincided with the detection of killer whale eDNA, and therefore could be attributed to predation.

The remaining 19 detections (68%) may be attributed to whale poop; while sinking rates and depths of baleen whale poop are generally unknown, we know from expert observations that fecal material from baleen whales is quite large and can sink rapidly from the surface (WHO CAN I PERS COMM FOR THIS?!?). Larger clumps of eDNA can be detectable for up to 2-4 days at the surface (CITE Pedro's paper here? and/or a Mod 1 paper?), and likely longer at depth where it is colder, darker, and there are fewer microbes digesting cellular material. Research on oceanic particle sinking rate indicates that smaller particles (up to 6 mm) can sink at rates of \~50-300 m day, and that sinking rates increase rapidly with particle size while degradation rates decrease with particle size, with maximum sinking depths in the range of \~800-5,000 m [@giering_etal20; @mcdonnell_etal15; @iversen_ploug10].

Based on this, the maximum sinking depth of larger poop particles may far exceed the z range of our study area. It is possible that baleen whale detection rates will continue to increase with increasing depth in some regions due to rapid particle sinking and longer eDNA residency times at depth (REF?). Similarly, detections made near the ocean floor that may be whalefall could also be signals of accumulated whale poop. Genearlly, eDNA is thought to travel less than 10 horizontal km before it is no longer detectable; therefore on a broad scale these detections are likely to reflect the xy distribution of these species but the temporal resolution of detections may be decreased from days to months in colder environments.

Beaked whales were also generally detected deeper than where they are thought to spend most of their time (@fig-H1POD-species), although still within their normal range of vertical habitat use. Beaked whales spend less of their time near the surface than either baleen whales or delphinids; time-at-depth models estimate that they spend approximately 70% of their time below 100 m, with depth ranges spanning down as far as 3,000 m (CITE Greg paper). While below the surface, beaked whales generally make deep (500+ m) foraging dives interspersed with several shallow (200-300 m) "resting" dives. We found that goose-beaked whales, considered the most common beaked whale in the Pacific Ocean, were most likely to be detected between 200 and 300 m, in the range of their resting dives. Baird's beaked whales were most likely to be detected at 400-500 m; as 500 m was the extent of our sampling range, it is possible that this species or other beaked whales have higher detection probabilities at depths greater than 500 m. While we did not make predictions of POD by depth for the saber-toothed whale, all three detections of this species were at 300 m (n = 1) and 500 m (n =2).

Finally, for both baleen whales and beaked whales, it is possible that the increase in detections at depth is related to the decreased abundance of competing organisms in the sample. Within each water sample that is collected, cetacean DNA is rare compared to that of microbes, plankton, fishes, etc (CITE from Nastassia's intro) and may be easily outcompeted for sequencing space in the lab. Deeper in the water column, where microbes, plankton, and fish DNA are likely to be less dense, their masking effect may be lessened enough to decrease the rate of false negatives. **We may be able to see this if detections at depth show up in more of the technical replicates than near the surface.**

In general, sampling near the surface will increase detection rates of delphinids (and likely porpoises), and sampling at 300-500 m or deeper will increase detection rates of baleen whales and beaked whales. While we found that POD varies by species, we also observed significant trends in POD with depth when species were group by phylogenetic Family or by broad prey category. This indicates some degree of similarity in habitat use across the z-axis within these groups, such that researchers wanting to design an eDNA monitoring study targeting a species for which there is little informaton could use Family or preferred prey to guide their decisions regarding sampling depth.

#### Question 2:

5.  Occupancy models can be used to separate effects of eDNA presence at the sampling site/in the bottle and detection in the sequence data.

6.  Increasing primer specificity increases POD. POD per station/depth increases \~linearly with technical replication, except that increasing freeze/thaw cycles decreases POD and may counteract the effect of technical replication (for longer target regions, as with Dloop in this study). Shorter regions more robust to freeze/thaw but often have less specificity and lower POD, and/or lower variability and ability to classify to species. **Add here how many detections were made at the family level that were dropped.**

Patin et al. \[Patin2025\] used a Bayesian hierarchical occupancy model to characterize delphinid occupancy, capture, and detection across sampling methods and primers in a highly replicated survey in Southern California. The high level of replication (biological and technical) in that study allowed the authors to separately estimate occupancy probability by depth, capture probability by sampling method, and detection probability by primer. Here, we had access to a single biological replicate collected at each sampling site and depth, so we are unable to resolve the difference between occupancy and capture of cetacean eDNA. However, in our study usually five and sometimes as many as eight technical replicates were run on each biological replicate, so we did use this information to better characterize occupancy/capture by depth and detectability by primer.

We found that species-specific patterns of cetacean eDNA presence with depth were qualitatively similar in shape when fitted in an occupancy model context, but the values tended to be much higher, with confidence intervals approaching 1. This makes sense since the Bayesian hierarchical occupancy model was able to separately estimate the probability of presence and the probability of detection, whereas the frequentist GAM was not.

#### Question 3:

Trends in POD by depth varied geographically by species, as illustrated by the three example species for which we made predictions of depth of maximum POD (@fig-H3PODxy). At the broad scale of this study, we expect that this at least partially reflects geographic variability in species-specific use of the water column. That may reflect geographic shifts in the depth distribution of prey or preferred prey items of a given predator, which may in turn reflect geographic variability in oceanographic conditions.

Alternatively, it is likely that at least a portion of this observed variability is directly linked to oceanographic variability, and its effect on the fate and transport of eDNA once deposited in the water column. eDNA degradation and residence time will be affected by water temperature, pH, and microbial density. During that time, horizontal transport will be affected by currents, eddies, etc. Vertical transport will be affected by upwelling, mixed layer depth, etc. This study is relatively insensitive to horizontal transport of eDNA, which is expected to travel up to XX km on average before it is no longer detectable (CITE Mod 1 paper here?), our samples were collected at 10 km intervals throughout the study area. Our understanding of the vertical distribution of eDNA is more likely to be affected by these oceanographic processes, which can change sinking rates or residence time.

In both cases, oceanography is an important driver of the 3D distribution of marine mammal eDNA; therefore our next study will focus on the development of 3D spatial distribution models using eDNA and using these models to determine the important oceanographic drivers of marine mammal eDNA in 3D space.

#### Marine Mammal monitoring: lessons for eDNA experimental design

1.  Sample at depth where species spends most of its time. If this is unknown, can use broad prey category or taxonomic group as a guide.

2.  In laboratory processing, sequence longer fragments first, using as few freeze/thaw cycles as possible. Increasing primer specificity and technical replicates increases probability of detecting a species at a diven location/depth. Shorter fragments that were insensitive to freeze/thaw may be a useful way to increase technical replication but will require more technical replicates than the longer and more specific target regions to acheive the same probability of detection.

3.  We didn't explicity consider variability in shedding, dispersal, or current in this study, which may affect downstream interpretation of results.

4.  Increasing spatiotemporal sampling resolution could provide insight about 3D habitat use.

#### Conclusions section?

## Data availability

All raw sequence data have been uploaded to GenBank (Accessions: XXXX). Detection data, sample metadata, and all code used to generate this manuscript are available at https://github.com/MMARINeDNA/zDistribution.

## Acknowledgements

We are grateful to the many people who supported eDNA sample collection and processing, including XXXX, and the crew of the R/V XXXX. Funding for this research came from ONR Grant XXXX.

## Supplemental Materials

Table S1. Sample metadata.

```{r}
#| label: tbl-m12AUC
#| eval: FALSE
#| tbl-cap: "Table S2. Per-species AUC, optimized TSS, and optimized TSS threshold for Q1 and Q3 models."

# m3.0c_metrics <- AUC_df_m3.0c %>% 
#   summarize(AUC = mean(AUC)) %>% 
#   bind_cols(TSS_df_m3.0c %>% arrange(desc(mean)) %>% slice_head()) %>% 
#   select(AUC, threshold, mean) %>% 
#   rename("TSS" = mean, "Threshold" = threshold) %>% 
#   mutate(species_model = "Q3")
# 
# model_metrics <- species_metrics %>% 
#   mutate(species_model = paste0(species, "-Q1")) %>% 
#   select(-species) %>% 
#   rename("AUC" = meanAUC, "Threshold" = meanThreshold, "TSS" = meanTSS) %>% 
#   bind_rows(m3.0c_metrics)
# 
# pander(m3.0c_metrics)
```

```{r}
#| label: fig-S1-null
#| fig-cap: "Supplemental Figure S1. Q1 null hypothesis plot."

m1.0POD

```

```{r}
#| label: fig-S2-TSS
#| fig-cap: "Supplemental Figure S2. TSS plots for the best-fit Q1 model (top) and Q3 model (bottom) showing the 10-fold mean TSS values across all detection thresholds. The optimal detection threshold is that which maximizes TSS."

TSS_plot / TSS_plot_m3.0c

```

```{r}
#| label: fig-S3-whalefall
#| fig-cap: "Supplemental Figure S3. Locations of eDNA detections within 100 m of the ocean floor, which may represent nearby whalefall."

whalefall_map
```

```{r}
#| label: fig-SuppH3PODxy_baleen
#| eval: FALSE
#| fig-cap: "Supplemental Figure S4. Depth of maximum probability of detection (POD; top) and range of depths falling within 50% CI of the depth of maximum detection (bottom) for humpback whales, separating likely whale poop (detections > 200 m) from detections within the normal depth range of humpback whales."
#| fig-width: 7
#| fig-height: 15

#depthPODmax_baleen

```

```{r}
#| label: fig-SuppH1POD_obs
#| fig-cap: "Supplemental Figure S5. Q1 best fit model with observed detections."
#| fig-width: 15
#| fig-height: 7

m1.2PODwObs

```
