---
format: html
editor: visual
---

We selected `r length(unique(detect_data_clean$NWFSCsampleID))` water samples collected at `r length(unique(detect_data_clean$station))` unique locations during the 2019 NOAA hake abundance survey (@fig-bubble), targeting surface, 50, 150, 300, and 500 m sampling depths at each station. We amplified and sequenced these samples in 1 technical replicate using MFU, 3 technical replicates using MV1, and 1 technical replicate using DL, resulting in a final total of `r detect_data_clean %>% group_by(primer, NWFSCsampleID) %>% n_groups()` successfully amplified and sequenced samples.

```{r}
#| label: tbl-sp-detects
#| tbl-cap: "Scientific names, common names, taxonomic family, broad prey category, and detections for each cetacean species detected in our study area. Detections are split by primer, and MV1 detections are split by whether a species was detected in 1, 2, or 3 of the 3 total technical replicates for that primer."

table <- mmEcoEvo %>% 
  select(Species, common_name, Family, Broad_taxa, Prey.family) %>% 
  inner_join(detect_primer_techRep_species_clean, by = c("Species" = "BestTaxon")) %>% 
  rename("Common name" = common_name, "Prey family" = Prey.family, "Broad taxon" = Broad_taxa) %>% 
  #left_join(detect_per_primer_species_clean, by = c("Species" = "BestTaxon")) %>%
  mutate(`Broad taxon` = tolower(`Broad taxon`))

pander(table, split.tables = Inf)
```

Across all samples, we observed `r sum(detect_data_clean$Detected)` positive detections of `r length(unique(detect_data_clean$BestTaxon))` cetacean species, including 6 baleen whale species, 5 dolphin species, 2 porpoise species, and 3 beaked whale species (@tbl-sp-detects). `r nrow(detect_per_species_clean %>% filter(nDetect > 10))` species were observed more than 10 times, and and additional `r nrow(detect_per_species_clean %>% filter(nDetect < 10))` were observed fewer than 10 times. We also detected bottlenose dolphins (*Tursiops truncatus*) a total of one time in the dataset; this detection occurred north of the known range of this species in the California Current [@carretta_etal24]. As noted in the methods section, we consider this detection likely to have resulted from lab contamination and removed it before downstream processing.

@fig-bubble illustrates the distribution of detections across sampling depths $z$, ignoring $xy$ distribution, for all cetacean species included in the study. Notably, our dataset includes one unexpected detection of a bowhead whale (*Balaena mysticetus*), not generally observed south of 63 N latitude in the Bering Sea since the end of commercial whaling and not known to ever inhabit the California Current [@rugh_etal03], as well as three detections of extremely rare saber-toothed whales (*Mesoplodon stejnegeri*), a cold-water beaked whale observed only twice, in the Gulf of Alaska, in 1986-2005 [@Hamilton2009a]. This is the first direct observation of this species in the California Current to date; it previously was only known to inhabit the region based on two stranding events from May 2015 in Oregon and Washington [@Hamilton2009a; @moore_barlow17; @Becker2020].

```{r}
#| label: fig-bubble
#| fig-cap: "Left: Distribution of detections across sampling depths for all species included in the study. Species are grouped as follows: baleen whales includes species in Mysticeti (Balaenopteridae, Balaenidae, and Eschericthidae); dolphins/porpoises includes species in Delphinoidea (Delphinidae and Phocoenidae); beaked whales includes species in Ziphiidae. Right: Geographic distribution of samples used in this study, acquired from water collected during a 2019 survey of hake distribution in the California Current. "
#| fig-width: 8
#| fig-height: 6

aligned <- align_plots(detectDepth_bubble, map, align = "v", axis = "tblr")
plot_grid(plotlist = aligned, ncol = 2, rel_widths = c(3, 1))
```

#### Question 1: Quantifying the effect of observation bias on detectability

Given the presence of cetacean eDNA in a sample, its POD is affected by primer choice (@fig-H2POD-obsbias, left). Assuming a single freeze/thaw cycle and technical replicate, MV1 had the highest expected POD ($\Omega_\text{m*t}$, `r round(mean(thaw1_pdetect %>% filter(Primer == "MV1") %>% pull("PDetect")), digits = 2)`), followed by DL (`r round(mean(thaw1_pdetect %>% filter(Primer == "DL1") %>% pull("PDetect")), digits = 2)`) and then MFU (`r round(mean(thaw1_pdetect %>% filter(Primer == "MFU") %>% pull("PDetect")), digits = 2)`). Using the mean posterior POD for each primer at one freeze/thaw cycle ($\Omega_{m*1}$), we calculated the expected rate of increase in POD with increasing technical replication by primer (@fig-H2POD-obsbias, middle). Based on our results, achieving a POD \>90% for cetacean DNA collected in a sample would require four technical replicates of MV1, \~5-6 technical replicates of DL, or seven technical replicates of MFU.

While increasing the number of technical replicates increases POD, the number of freeze/thaw cycles decreases POD (@fig-H2POD-obsbias, right), consistent with increased degradation with increasing freeze/thaw cycles. We found a relationship between freeze/thaw cycle and posterior estimates of POD for the longest primer in our study (DL, 360 bp), and a weaker relationship between freeze/thaw cycle and posterior POD for MV1 (202 bp). There was no relationship detected between posterior POD and freeze/thaw cycles for MFU (185 bp), noting limited statistical power to detect the effect of freeze/thaw as only 5 of 516 total MFU samples sequenced in this study had \> 2 thaw cycles.

Given our posterior estimates of POD by primer and number of freeze/thaw cycles, we calculated a sample-specific POD for each sample in our study, assuming cetacean DNA is present in the biological sample. Averaged across all samples in this study when accounting for freeze/thaw cycles, the mean detectability for cetacean DNA in a single technical replicate of DL is 0.14, for MFU is 0.31, and for MV1 is 0.39. Despite the significant reduction in POD in our DL sequence with increasing degradation caused by increasing freeze/thaw cycles (lowest DL POD = 0.08), including an additional four replicates using primers that target shorter DNA fragments that were not as strongly affected by freeze/thaw cycle (MV1 and MFU) ensured a high overall POD for cetacean eDNA in a biological sample, conditional on its presence. We estimated the sample-specific predicted DNA POD for each sample in our study using the sample-specific parameters for number of freeze/thaw cycles and primer used in each technical replicate and calculating POD across all technical replicates for each sample combined. Using this approach, we determined that mean predicted POD across all samples in the study was `r round(mean(POD_per_samp$POD), digits = 2)` (range: `r round(min(POD_per_samp$POD), digits = 2)` - `r round(max(POD_per_samp$POD), digits = 2)`).

```{r}
#| label: fig-H2POD-obsbias
#| fig-cap: "Left: Posterior distribution of primer-specific POD (vertical axis) for cetacean eDNA given its presence in a sample (horizontal axis) assuming one freeze-thaw cycle. Middle: Posterior mean primer-specific POD (vertical axis) with increasing technical replication (horizontal axis) for each primer used in this study, also assuming one freeze/thaw cycle. Right: Posterior distribution of POD given presence (vertical axis) by number of freeze/thaw cycles (horizontal axis) and primer (panels)."
#| fig-width: 10
#| fig-height: 6

 pdetect_primer | Q2_TechRepsXPrimer | ppdetect

```

### Question 2: Does POD vary with depth?

In the next two sections we switch to considering detection rate of cetacean DNA in the ocean using five technical replicates per sample (3 MV1, 1 DL, 1 MFU), without assuming its presence in a biological sample. Detection rate varies with depth (p \<\< 0.001, deviance explained = 0.7%; Supplemental Figure S1); the model that best describes variance in detection rate with depth includes significant smooth terms for depth, species, and a depth x species interaction (AIC = `r round(min(modelAIC$AIC))`, @tbl-modelAIC). This model explained 14.3% of variance in eDNA detection probability, with a mean AUC of `r round(mean(AUC_df_clean$AUC), digits = 2)` and a mean optimized TSS of `r round(max(TSS_df_clean$mean), digits = 2)`. Per-species AUC and TSS values are shown in Supplemental Table S3 and Supplemental Figure S2. Finally, we found general agreement between species-specific model predictions and raw detection data via visual comparison (Supplemental Figure S3).

```{r}
#| label: tbl-modelAIC
#| tbl-cap: "Akaike Information Criterion values for all GAMs tested to characterize the relationship between depth and detection rate of eDNA. Models and stratification schemes from Eqns. 4-6."

modelAICtable <- modelAIC %>% 
  arrange(AIC) %>% 
  mutate(minAIC = min(AIC)) %>% 
  mutate("$\u394 AIC$" = AIC-minAIC) %>% 
  mutate("Stratification" = c("species", "species", "family", "prey", "none")) %>% 
  mutate(Model = case_when(model == "m1.2"~"$\u03b2_s + f_s(z_i)$ (Eq. 6)",
                           model == "m1.1"~"$\u03b2_s + f(z_i)$ (Eq. 5)",
                           model == "m1.2a"~"$\u03b2_s + f_F(z_i)$ (Eq. 7)",
                           model == "m1.2b"~"$\u03b2_s + f_e(z_i)$ (Eq. 8)",
                           model == "m1.0"~"$\u03b2_0 + f(z_i)$ (Eq. 4)", 
                           TRUE~model)) %>% 
  mutate("$df$" = df) %>% 
  select(Model, Stratification, "$df$", "$\u394 AIC$")

kable(modelAICtable, escape = FALSE, digits = 1)
```

Broadly, delphinid detection rate was highest at the surface, beaked whale detection rate was highest at depths \> 200 m, and baleen whale detection rate was highest either at the surface or at depths \> 400 m (@fig-H1POD-species). Note that although detection data from the five species detected fewer than 10 times were included in our model input, we did not make predictions for any of these species.

Among the delphinids, species-specific trends in detection rate generally reflected previously developed estimates of time-at-depth [@watwood_buonantony12; @u_s_departmentofthenavy17]. Conversely, most beaked whale and baleen whale species in the study were very likely to be detected outside depths where they were estimated to spend most of their time (@fig-H1POD-species). Most notably, predicted detection rate depth distribuitons for two of the three baleen whales indicated a high detection rate at 500 m, while these two species spend most of their time above \~200 m depth. Per-species model predictions are shown alongside observed detections in Supplemental Figure S5.

```{r}
#| label: fig-H1POD-species
#| fig-cap: "Species-specific detection rate predictions across all depths sampled in this study. Solid line shows posterior predicted detection rates; shaded regions show 95%, 75%, and 50% confidence band. Dashed line shows proportional time at depth. Y-axis scale is the same for detection rate (solid line) and proportional time at depth (dashed line). Plots are colored by broad taxonomic category as described in Table 1."
#| fig-width: 10
#| fig-height: 6

m1.2POD

```

Models stratifying detections by taxonomic family and preferred prey type both had lower AIC values than the model stratified by species (AIC = `r round(filter(modelAIC, model == "m1.2a")$AIC, digits = 0)` and `r round(filter(modelAIC, model == "m1.2b")$AIC, digits = 0)`, respectively); however, these models still had significant terms for family and depth-by-family interactions (taxonomic model) and for depth, prey type, and depth-by-prey type interactions (prey type model). Predicted detection rate curves for each model illustrate varying trends in detection rate with depth across strata (@fig-H1POD-family-prey). Similar to the model stratified by species, we did not make predictions for families with fewer than 10 detections using the taxonomic model, which excluded the Balaenidae (bowhead whale family).

```{r}
#| label: fig-H1POD-family-prey
#| fig-cap: "Left: Detection rate by depth for species stratified by taxonomic family. Right: Detection rate by depth for species stratified based on broad prey categories."
#| fig-width: 10
#| fig-height: 6

m1.2aPOD_bordered <- patchwork::wrap_elements(m1.2aPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

m1.2bPOD_bordered <- patchwork::wrap_elements(m1.2bPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

(m1.2aPOD_bordered + m1.2bPOD_bordered) +
  plot_layout(widths = c(3, 1), axis_titles = "collect") +
  theme(legend.position = "bottom")

```

#### Question 3: Does depth variability in POD covary with geographic variability?

Adding covariates for latitude ($y$) and longitude ($x$) to the best fit model from Q2 improved overall model performance. Our model included terms for species, lat/lon, and depth, as well as interaction terms for species-by-lat/lon, species-by-depth, and species-by-depth-by-lat/lon (AIC = `r round(AIC(m3.0c_clean), digits = 0)`), and explained 25.3% of the variability in eDNA detection rate. Although terms for depth and lat/lon were not significant, removing these terms increased model AIC to 4947. Mean AUC across a 10-fold validation test was `r round(mean(AUC_df_m3.0c_clean$AUC), digits = 2)` and mean optimized TSS was `r round(max(TSS_df_m3.0c_clean$mean), digits = 2)`. Full model TSS for all thresholds are shown in Supplemental Table S3 and Supplemental Figure S2.

Although $xy$ improved model performance, the predicted $z$ distribution of detection rate for the most frequently detected delphinid (Pacific white-sided dolphin, *Aethalodelphis obliquidens*), baleen whale (humpback whale, *Megaptera novaeangliae*), and beaked whale (northern giant bottlenose whale, *Berardius bairdii*) exhibited only marginal variability across $x$ and $y$, largely adhering to the broad trends identified in Q2 (@fig-H3PODxy).

The Pacific white-side dolphin detection rate is maximized near the surface throughout the study area, with maximum detection rate between depths of `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 350) %>% pull(depth), na.rm = TRUE)` m. The depth that maximizes detection rate increased toward the southeast corner of the study area. Detection rate was less than 1% in the region off the shelf break between Point Conception and San Francisco Bay.

Depths that maximized detection rate for northern giant bottlenose whales varied between `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m, with no observable trends in spatial variability in across the study area. Notably, detection rate was less than 1% along most of the continental shelf.

Depths that maximized detection rate for humpback whales had two disparate peaks, at `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m, with almost no intermediate depths that maximized detection rate. The two peaks were roughly distributed so that detection rate was maximized at the surface along the continental shelf, and maximized at depth on the shelf break or in deeper water. Detection rate was less than 1% along the coast near Point Conception, as well as off the shelf in the same region.

```{r}
#| label: fig-H3PODxy
#| fig-cap: "Top: Depths that maximize detection rates for three species throughout the study area: northern giant bottlenose whale, max detecton rate ~ 0.3; Pacific white-sided dolphin, max detection rate ~ 0.35; humpback whale, max detection rate ~ 0.54. Observed detection depths are shown as colored points. Bottom: Model uncertainty represented as the strip width of depths within the 50% CI for maximum detecion rate of the same three species. Regions with detection rate < 0.01 throughout the water column were masked in light grey; 500 m, 1,000 m, and 2,000 m contour lines are shown in dark grey."
#| fig-width: 6
#| fig-height: 20

depth_max_detect_grid

# (depth_max_detect_clean | ci95_plot_clean) + 
#   plot_layout(guides = "collect") &
#   theme(legend.position = "right")

```

In addition to examining geographic patterns in the depths that maximize detection rate, we examined uncertainty around these values by determining the strip width of depths that fell within the 50% confidence interval for predicted maximum detecion rate (@fig-H3PODxy, bottom). In the northern portion of the study area, where Pacific white-sided dolphins are most likely to be observed near the surface, uncertainty around depth maximum detection rate is negligible (minimum range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depthWidth), na.rm = TRUE)` m wide). That uncertainty increases toward the south and east (maximum range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 400) %>% pull(depthWidth), na.rm = TRUE)` m wide), with the greatest uncertainty near the region where detection rate is close to 0. Uncertainty around the depths maximizing detection rate for northern giant bottlenose whales was minimal in most of the study area (minimum range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m wide), except for a small portion of the northwestern edge of the study area (maximum range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m wide).

The bimodal peak in depths that maximize detection rate that we observed in humpback whale eDNA was mirrored in the patterns we observed in uncertainty: the depth range was largest in the transition zones between the two peaks (maximum range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m wide), and small outside of the transition zones (minimum range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m wide).
