---
format: html
editor: visual
---

We selected `r length(unique(detect_data_clean$NWFSCsampleID))` water samples collected at `r length(unique(detect_data_clean$station))` unique locations during the 2019 NOAA hake abundance survey (@fig-bubble), targeting 0, 50, 150, 300, and 500 m sampling depths at each station. Where the bottom was less than 500 m deep, we did not collect a sample at 500 m. We amplified and sequenced these samples in 1 technical replicate using MFU, 3 technical replicates using MV1, and 1 technical replicate using DL, resulting in a final total of `r detect_data_clean %>% group_by(primer, NWFSCsampleID) %>% n_groups()` amplified and sequenced samples.

```{r}
#| label: tbl-sp-detects
#| tbl-cap: "Scientific names, common names, abbrevation, taxonomic family, broad prey category, and total number of detections for each cetacean species detected in our study area."

table <- mmEcoEvo %>% 
  select(Species, common_name, abbrev, Family, Broad_taxa, Prey.family) %>% 
  inner_join(detect_per_species_clean, by = c("Species" = "BestTaxon")) %>% 
  arrange(desc(nDetect)) %>% 
  rename("# Detections" = nDetect, "Common name" = common_name, "Sp. abbrevation" = abbrev, "Prey family" = Prey.family, "Broad taxon" = Broad_taxa) %>% 
  left_join(detect_per_primer_species_clean, by = c("Species" = "BestTaxon"))

pander(table)
```

Across all samples, we observed `r sum(detect_data_clean$Detected)` positive detections of `r length(unique(detect_data_clean$BestTaxon))` cetacean species, including 6 baleen whale species, 5 dolphin species, 2 porpoise species, and 3 beaked whale species (@tbl-sp-detects). `r nrow(detect_per_species_clean %>% filter(nDetect > 10))` species were observed more than 10 times, and and additional `r nrow(detect_per_species_clean %>% filter(nDetect < 10))` were observed fewer than 10 times. This includes one detection of a bowhead whale (*Balaena mysticetus*), as well as three detections of saber-toothed whales (*Mesoplodon stejnegeri*). @fig-bubble illustrates the distribution of detections across sampling depths z, ignoring xy distribution, for all cetacean species included in the study.

```{r}
#| label: fig-bubble
#| fig-cap: "Left: Distribution of detections across sampling depths for all species included in the study. Species are grouped as follows: Baleen whales includes the Balaenopteridae, Balaenidae, and Eschericthidae; Dolphins/Porpoises includes all Delphinidae and Phocoenidae; Beaked whales includes all Ziphiidae. Right: Geographic distribution of samples used in this study, acquired from water collected during a 2019 survey of hake distribution in the California Current. "
#| fig-width: 8
#| fig-height: 6

aligned <- align_plots(detectDepth_bubble, map, align = "v", axis = "tblr")
plot_grid(plotlist = aligned, ncol = 2, rel_widths = c(3, 1))
```

#### Question 1: Quantifying the effect of observation bias on POD

Given the presence of cetacean eDNA in a sample, the probability of detecting that DNA is affected by primer specificity. On average across all species and samples, MV1 had the highest predicted POD (XX%), followed by MFU (XX%) and then DL (XX%). Therefore, the rate of increase in POD with technical replication also varies by primer (@fig-H2POD-obsbias). For all three primers, achieving a POD of \>90% given the presence of cetacean DNA in the sample would require a minimum of six technical replicates.

While increasing the number of technical replicates will increase POD, DNA degradation by freeze/thaw cycles decreased the probability of detecting cetacean DNA in a given sample (@fig-H2POD-obsbias). DL and MV1 had similar predicted POD at thaw = 1 (\~0.25), but predicted POD for DL declined to 0 with increasing freeze/thaw cycles while predicted POD for MV1 declined to \~0.13. The POD for MFU at thaw = 1 was lower than DL and MV1 (\~0.125). There was no significant relationship detected between predicted POD for MFU and freeze/thaw cycles, noting that only 5 of 516 total MFU samples had \> 2 thaw cycles.

Given a per station/depth coverage of 3 MV1, 1 MFU, and 1 DL technical replicates, we estimate a mean POD for each station/depth sample of 0.79 (upper CI = 1, lower CI = 0.37). **Note that these numbers are estimated from the predictions from the GLM but that model does not condition on presence of DNA in the sample. Not sure if it would be better to use numbers from the nimble model.**

```{r}
#| label: fig-H2POD-obsbias
#| fig-cap: "Left: Primer-specific detection probability with increasing technical replication for each primer used in this study. MFU = Mifish, MV1 = MarVer1, DL = Dloop. Right: Primer-specific effect of increasing number of freeze/thaw cycles."
#| fig-width: 10
#| fig-height: 6

 Q2_TechRepsXPrimer | PODthaw_plot

 #Replace TechRepsXPrimer with plot excluding DL1/2 split
 #Add Q2_detectability once it's rerun in the new ggplot2

```

### Question 2: Does POD vary with depth?

We first negated the null hypothesis of no variation across depth (Supplemental Figure S1), then determined that the best alternative model includes significant smooth terms for depth, species, and depth x species interaction (AIC = `r round(min(modelAIC$AIC))`, @tbl-modelAIC). This model explained 14.3% of variance in eDNA detection probability, with a mean AUC of `r round(mean(AUC_df_clean$AUC), digits = 2)` and a mean optimized TSS of `r round(max(TSS_df_clean$mean), digits = 2)`. Full model TSS for all thresholds, as well as per-species AUC, optimized TSS, TSS values across all thresholds are shown in Supplemental Table S2 and Supplemental Figure S2. Species-specific model predictions generally agreed with the raw detection data.

```{r}
#| label: tbl-modelAIC
#| tbl-cap: "Akaike Information Criterion values for all GAMs tested to characterize the relationship between depth and POD of eDNA."

modelAICtable <- modelAIC %>% 
  arrange(AIC) %>% 
  mutate(minAIC = min(AIC)) %>% 
  mutate(deltaAIC = AIC-minAIC) %>% 
  mutate("stratification" = c("species", "species", "family", "prey", "none")) %>% 
  mutate(model = case_when(model == "m1.2"~"depth, s(species)",
                           model == "m1.1"~"depth + species",
                           model == "m1.2a"~"depth, s(family)",
                           model == "m1.2b"~"depth, s(prey type)",
                           model == "m1.0"~"depth", 
                           TRUE~model)) %>% 
  select(model, stratification, df, deltaAIC)

pander(modelAICtable)
```

Broadly, delphinids were more likely to be detected at the surface, beaked whales at depths \> 200m, and baleen whales either at the surface or at depths \> 400m (@fig-H1POD-species). Note that although detection data from the five species detected fewer than 10 times were included in our model input, we did not make predictions for any of these species.

Among the delphinids, species-specific trends in POD generally reflected model estimates of time-at-depth. Conversely, most beaked whale and baleen whale species in the study were very likely to be detected outside depths where they spend most of their time (@fig-H1POD-species). Most notably, predicted POD-at-depth plots for two of the three baleen whales indicated a high probability of detection at 500 m, while these two species spend most of their time above \~200 m. Per-species model predictions are shown alongside observed detections in Supplemental Figure S5.

```{r}
#| label: fig-H1POD-species
#| fig-cap: "Species-specific probability of detection across all depths sampled in this study. Solid line shows POD; shaded regions show 95%, 75%, and 50% confidence intervals. Dashed line shows proportional time at depth. Species abbreviation codes can be found in Table 1; plots are colored by taxonomic category as outlined in Table 1."
#| fig-width: 10
#| fig-height: 6

m1.2POD

```

Models stratifying detections by taxonomic Family and preferred prey type both had lower AIC values than the model stratified by species (AIC = `r round(filter(modelAIC, model == "m1.2a")$AIC, digits = 0)` and `r round(filter(modelAIC, model == "m1.2b")$AIC, digits = 0)`, respectively); however, these models still had significant terms for Family and depth x Family interactions (taxonomic model) and for depth, prey type, and depth x prey type interactions (prey type model). Predicted POD curves for each model illustrate varying trends in POD with depth across strata (@fig-H1POD-family-prey). Similar to the model stratified by species, we did not make predictions for Families with fewer than 10 detections using the taxonomic model, which excluded the Balaenidae (bowhead whale family).

```{r}
#| label: fig-H1POD-family-prey
#| fig-cap: "Left: POD by depth for species stratified by taxonomic Family. Right: POD by depth for species stratified based on broad prey categories."
#| fig-width: 10
#| fig-height: 6

m1.2aPOD_bordered <- patchwork::wrap_elements(m1.2aPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

m1.2bPOD_bordered <- patchwork::wrap_elements(m1.2bPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

(m1.2aPOD_bordered + m1.2bPOD_bordered) +
  plot_layout(widths = c(3, 1), axis_titles = "collect") +
  theme(legend.position = "bottom")

```

#### Question 3: Does depth variability in POD covary with geographic variability?

Adding xy covariates to the best fit model from Q2 improved overall model performance. Our model included terms for species, lat/lon, and depth, as well as species x lat/lon, species x depth, and species x depth x lat/lon (AIC = `r AIC(m3.0c_clean)`), and explained 25.3% of the variability in eDNA detection. Although terms for depth and lat/lon were not significant, removing these terms increased model AIC to 4947. Mean AUC across a 10-fold validation test was `r round(mean(AUC_df_m3.0c_clean$AUC), digits = 2)` and mean optimized TSS was `r round(max(TSS_df_m3.0c_clean$mean), digits = 2)`. Full model TSS for all thresholds are shown in Supplemental Table S2 and Supplemental Figure S2.

Although xy predictor variables improved model performance, the predicted vertical probability of detection for the most frequently detected delphinid, baleen whale, and beaked whale exhibited minimal variability with latitute and longitude, largely adhering to the broad trends identified in Q2 (@fig-H3PODxy).

The Pacific white-side dolphin is most likely to be detected near the surface, with max POD varying between depths of `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 400) %>% pull(depth), na.rm = TRUE)` m, and deeper max POD depths toward the southeast corner of the study area. Depths of highest POD for Baird's beaked whales varied between `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m, with minimal spatial trends in this variability across the study area.

Depths of highest POD for humpback whales had two disparate peaks, at `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m, with no intermediate depths at max POD. The two peaks in depth of greatest POD for humpback whales were roughly distributed so that higher probabilities of detection at the surface occurred in patches along the continental shelf, and higher probabilities of detection at depth occured on the shelf break or in deeper water.

```{r}
#| label: fig-H3PODxy
#| fig-cap: "Depth of maximum probability of detection (POD; top) and range of depths falling within 50% CI of the depth of maximum detection (bottom) for three species: Baird's beaked whale (*B. bairdii*), max POD ~ 0.3; Pacific white-sided dolphin (*L. obliquidens*), max POD ~ 0.35; humpback whale (*M. novaegliae*), max POD ~ 0.54. Regions with max POD < 0.02 were masked in light grey."
#| fig-width: 7
#| fig-height: 15

depth_max_detect_clean / ci95_plot_clean

```

In addition to examining geographic patterns in the depth of maximum POD, we examined uncertainty around these values by mapping the range of depths that fell within the 50% confidence interval of maximum POD (@fig-H3PODxy, bottom). In the northern portion of the study area, where Pacific white-sided dolphins are most likely to be observed near the surface, the range of depths within the 50% CI of maximum POD is small (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depthWidth), na.rm = TRUE)` m). That range increases toward the south and east (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 400) %>% pull(depthWidth), na.rm = TRUE)` m), as the depth of maximum POD gets deeper for this species. Baird's beaked whales exhibited minimal minimal variability in 50% CI width of maximum depth (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m), except for a small portion of the northwestern edge of the study area (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m). The bimodal peak of maximum POD observed in humpback whale eDNA was mirrored in the range of depths within the 50% CI of maximum POD: the depth range was largest in the transition zones between the two peaks (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m), and small outside of the transition zones (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m).
