---
format: html
editor: visual
---

We selected `r length(unique(detect_data_clean$NWFSCsampleID))` water samples collected at `r length(unique(detect_data_clean$station))` unique locations during the 2019 NOAA hake abundance survey (@fig-bubble), targeting surface, 50, 150, 300, and 500 m sampling depths at each station. We amplified and sequenced these samples in 1 technical replicate using MFU, 3 technical replicates using MV1, and 1 technical replicate using DL, resulting in a final total of `r detect_data_clean %>% group_by(primer, NWFSCsampleID) %>% n_groups()` successfully amplified and sequenced samples.

```{r}
#| label: tbl-sp-detects
#| tbl-cap: "Scientific names, common names, abbreviation, taxonomic family, broad prey category, and total number of detections for each cetacean species detected in our study area."

table <- mmEcoEvo %>% 
  select(Species, common_name, abbrev, Family, Broad_taxa, Prey.family) %>% 
  inner_join(detect_per_species_clean, by = c("Species" = "BestTaxon")) %>% 
  arrange(desc(nDetect)) %>% 
  rename("# Detections" = nDetect, "Common name" = common_name, "Sp. abbrevation" = abbrev, "Prey family" = Prey.family, "Broad taxon" = Broad_taxa) %>% 
  left_join(detect_per_primer_species_clean, by = c("Species" = "BestTaxon")) %>%
  mutate(`Broad taxon` = tolower(`Broad taxon`))

pander(table, split.tables = Inf)
```

Across all samples, we observed `r sum(detect_data_clean$Detected)` positive detections of `r length(unique(detect_data_clean$BestTaxon))` cetacean species, including 6 baleen whale species, 5 dolphin species, 2 porpoise species, and 3 beaked whale species (@tbl-sp-detects). `r nrow(detect_per_species_clean %>% filter(nDetect > 10))` species were observed more than 10 times, and and additional `r nrow(detect_per_species_clean %>% filter(nDetect < 10))` were observed fewer than 10 or fewer times. @fig-bubble illustrates the distribution of detections across sampling depths $z$, ignoring $xy$ distribution, for all cetacean species included in the study. Notably, our dataset includes one unexpected detection of a bowhead whale (*Balaena mysticetus*), not generally observed south of 63 N latitude in the Bering Sea since the end of commercial whaling and not known to ever inhabit the California Current \[\@rugh_etal03\], as well as three detections of extremely rare saber-toothed whales (*Mesoplodon stejnegeri*), a cold-water beaked whale observed only twice, in the Gulf of Alaska, in 1986-2005 \[\@Hamilton2009a\]. This is the first direct observation of this in the California Current to date; it previously was only known to inhabit the region based on two stranding events from May 2015 in Oregon and Washington \[\@Hamilton2009a; \@moore_barlow17; \@Becker2020\].

```{r}
#| label: fig-bubble
#| fig-cap: "Left: Distribution of detections across sampling depths for all species included in the study. Species are grouped as follows: baleen whales includes species in Mysticeti (Balaenopteridae, Balaenidae, and Eschericthidae); dolphins/porpoises includes species in Delphinoidea (Delphinidae and Phocoenidae); beaked whales includes species in Ziphiidae. Right: Geographic distribution of samples used in this study, acquired from water collected during a 2019 survey of hake distribution in the California Current. "
#| fig-width: 8
#| fig-height: 6

aligned <- align_plots(detectDepth_bubble, map, align = "v", axis = "tblr")
plot_grid(plotlist = aligned, ncol = 2, rel_widths = c(3, 1))
```

#### Question 1: Quantifying the effect of observation bias on POD

Given the presence of cetacean eDNA in a sample, the probability of detecting that DNA is affected by primer use (@fig-H2POD-obsbias, left). Assuming a single freeze/thaw cycle and technical replicate, MV1 had the highest predicted detectability (`r round(mean(thaw1_pdetect %>% filter(Primer == "MV1") %>% pull("PDetect")), digits = 2)`), followed by DL (`r round(mean(thaw1_pdetect %>% filter(Primer == "DL1") %>% pull("PDetect")), digits = 2)`) and then MFU (`r round(mean(thaw1_pdetect %>% filter(Primer == "MFU") %>% pull("PDetect")), digits = 2)`). Using the mean posterior detectability for each primer at one freeze/thaw cycle ($\Omega_{primer*1}$), we calculated the expected rate of increase in POD in at least one technical replicate with technical replication by primer (@fig-H2POD-obsbias, middle), assuming that each technical replicate is an independent Bernoulli trial. Based on our results, achieving a detectability of \>90% for cetacean DNA collected in a sample would require four technical replicates of MV1, \~5-6 technical replicates of DL, or seven technical replicates of MFU.

While increasing the number of technical replicates increases detectability, the number of freeze/thaw cycles decreases detectability (@fig-H2POD-obsbias, right), consistent with increased degradation with increasing freeze/thaw cycles. We found a relationship between freeze/thaw cycle and posterior estimates of detectability for the longest primer in our study (DL, 360 bp), and a weaker relationship between freeze/thaw cycle and posterior POD for MV1 (202 bp). There was no relationship detected between posterior POD and freeze/thaw cycles for MFU (185 bp), noting limited power to detect the effect of freeze/thaw as only 5 of 516 total MFU samples sequenced in this study had \> 2 thaw cycles.

Given our posterior estimates of detection probability by primer and number of freeze/thaw cycles, we calculated a sample-specific probability of detection for each sample in our study, assuming cetacean DNA is present in the biological sample. Averaged across all samples in this study when accounting for freeze/thaw cycles, the mean detectability for a single technical replicate of DL is 0.14, for MFU is 0.31, and for MV1 is 0.39. Despite the significant reduction in detectability in our DL samples with increasing number of freeze/thaw cycles (lowest DL detectability = 0.08), including four replicates using primers that were not strongly affected by freeze/thaw cycle (MV1 and MFU) ensured a high overall probability of detecting cetacean eDNA in a biological sample, conditional on its presence. We estimated the sample-specific predicted DNA detectability for each sample in the study using the number of freeze/thaw cycles for each combination of primer and technical replicate, and determined that mean predicted detectability across all samples in the study was `r round(mean(POD_per_samp$POD), digits = 2)` (range: `r round(min(POD_per_samp$POD), digits = 2)` - `r round(max(POD_per_samp$POD), digits = 2)`).

```{r}
#| label: fig-H2POD-obsbias
#| fig-cap: "Left: Posterior distribution of primer-specific probability of detecting cetacean eDNA given its presence in a sample (horizontal axis) with one freeze-thaw cycle. Middle: Posterior mean primer-specific detection probability (vertical axis) with increasing technical replication (horizontal axis) for each primer used in this study. Dashed line indicates arbitrary target threshold of 0.90. Right: Posterior distribution of the probability of detection given presence (vertical axis) by number of freeze/thaw cycles (horizontal axis) and primer (panels). For all plots, MFU = Mifish, MV1 = MarVer1, DL = Dloop."
#| fig-width: 10
#| fig-height: 6

 pdetect_primer | Q2_TechRepsXPrimer | ppdetect

```

### Question 2: Does POD vary with depth?

We first negated the null hypothesis of no variation across depth and that depth does not interact with primer or freeze/thaw cycle to affect POD (not sure if this is the best way to say this) (Supplemental Figure S1, **add Q2 model of depth POD here??**). We then determined that the best alternative model describing POD includes significant smooth terms for depth, species, and a depth x species interaction (AIC = `r round(min(modelAIC$AIC))`, @tbl-modelAIC). This model explained 14.3% of variance in eDNA detection probability, with a mean AUC of `r round(mean(AUC_df_clean$AUC), digits = 2)` and a mean optimized TSS of `r round(max(TSS_df_clean$mean), digits = 2)`. Full model TSS for all thresholds, as well as per-species AUC, optimized TSS, TSS values across all thresholds are shown in Supplemental Table S2 and Supplemental Figure S2. Species-specific model predictions generally agreed with the raw detection data.

```{r}
#| label: tbl-modelAIC
#| tbl-cap: "Akaike Information Criterion values for all GAMs tested to characterize the relationship between depth and POD of eDNA."

modelAICtable <- modelAIC %>% 
  arrange(AIC) %>% 
  mutate(minAIC = min(AIC)) %>% 
  mutate(deltaAIC = AIC-minAIC) %>% 
  mutate("stratification" = c("species", "species", "family", "prey", "none")) %>% 
  mutate(model = case_when(model == "m1.2"~"depth, s(species)",
                           model == "m1.1"~"depth + species",
                           model == "m1.2a"~"depth, s(family)",
                           model == "m1.2b"~"depth, s(prey type)",
                           model == "m1.0"~"depth", 
                           TRUE~model)) %>% 
  select(model, stratification, df, deltaAIC)

pander(modelAICtable)
```

Broadly, delphinids were more likely to be detected at the surface, beaked whales at depths \> 200m, and baleen whales either at the surface or at depths \> 400m (@fig-H1POD-species). Note that although detection data from the five species detected fewer than 10 times were included in our model input, we did not make predictions for any of these species.

Among the delphinids, species-specific trends in POD generally reflected model estimates of time-at-depth. Conversely, most beaked whale and baleen whale species in the study were very likely to be detected outside depths where they spend most of their time (@fig-H1POD-species). Most notably, predicted POD-at-depth plots for two of the three baleen whales indicated a high probability of detection at 500 m, while these two species spend most of their time above \~200 m. Per-species model predictions are shown alongside observed detections in Supplemental Figure S5.

```{r}
#| label: fig-H1POD-species
#| fig-cap: "Species-specific probability of detection across all depths sampled in this study. Solid line shows POD; shaded regions show 95%, 75%, and 50% confidence intervals. Dashed line shows proportional time at depth. Species abbreviation codes can be found in Table 1; plots are colored by taxonomic category as outlined in Table 1."
#| fig-width: 10
#| fig-height: 6

m1.2POD

```

Models stratifying detections by taxonomic Family and preferred prey type both had lower AIC values than the model stratified by species (AIC = `r round(filter(modelAIC, model == "m1.2a")$AIC, digits = 0)` and `r round(filter(modelAIC, model == "m1.2b")$AIC, digits = 0)`, respectively); however, these models still had significant terms for Family and depth x Family interactions (taxonomic model) and for depth, prey type, and depth x prey type interactions (prey type model). Predicted POD curves for each model illustrate varying trends in POD with depth across strata (@fig-H1POD-family-prey). Similar to the model stratified by species, we did not make predictions for Families with fewer than 10 detections using the taxonomic model, which excluded the Balaenidae (bowhead whale family).

```{r}
#| label: fig-H1POD-family-prey
#| fig-cap: "Left: POD by depth for species stratified by taxonomic Family. Right: POD by depth for species stratified based on broad prey categories."
#| fig-width: 10
#| fig-height: 6

m1.2aPOD_bordered <- patchwork::wrap_elements(m1.2aPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

m1.2bPOD_bordered <- patchwork::wrap_elements(m1.2bPOD) +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1))

(m1.2aPOD_bordered + m1.2bPOD_bordered) +
  plot_layout(widths = c(3, 1), axis_titles = "collect") +
  theme(legend.position = "bottom")

```

#### Question 3: Does depth variability in POD covary with geographic variability?

Adding xy covariates to the best fit model from Q2 improved overall model performance. Our model included terms for species, lat/lon, and depth, as well as species x lat/lon, species x depth, and species x depth x lat/lon (AIC = `r round(AIC(m3.0c_clean), digits = 0)`), and explained 25.3% of the variability in eDNA detection. Although terms for depth and lat/lon were not significant, removing these terms increased model AIC to 4947. Mean AUC across a 10-fold validation test was `r round(mean(AUC_df_m3.0c_clean$AUC), digits = 2)` and mean optimized TSS was `r round(max(TSS_df_m3.0c_clean$mean), digits = 2)`. Full model TSS for all thresholds are shown in Supplemental Table S2 and Supplemental Figure S2.

Although $xy$ predictor variables improved model performance, the predicted $z$ distribution of POD for the most frequently detected delphinid, baleen whale, and beaked whale exhibited minimal variability across $x$ and $y$, largely adhering to the broad trends identified in Q2 (@fig-H3PODxy).

The Pacific white-side dolphin is most likely to be detected near the surface, with max POD varying between depths of `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 350) %>% pull(depth), na.rm = TRUE)` m, and deeper max POD depths toward the southeast corner of the study area. POD was considered 0 (less than 1%) in the region off the shelf break between Point Conception and San Fransisco Bay (@fig-H3PODxy).

Depths of highest POD for Baird's beaked whales varied between `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depth), na.rm = TRUE)` m, with minimal spatial variability in depth of max POD across the study area. Notably, POD was considered 0 (less than 1%) along most of the continental shelf.

Depths of highest POD for humpback whales had two disparate peaks, at `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m and `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depth), na.rm = TRUE)` m, with almost no intermediate depths at max POD. The two peaks in depth of greatest POD for humpback whales were roughly distributed so that higher probabilities of detection at the surface occurred along the continental shelf, and higher probabilities of detection at depth occured on the shelf break or in deeper water. POD was considered 0 (less than 1%) along the coast near Point Conception, as well as off the shelf in the same region.

```{r}
#| label: fig-H3PODxy
#| fig-cap: "Depth of maximum probability of detection (POD; top) and range of depths falling within 50% CI of the depth of maximum detection (bottom) for three species: Baird's beaked whale (*B. bairdii*), max POD ~ 0.3; Pacific white-sided dolphin (*L. obliquidens*), max POD ~ 0.35; humpback whale (*M. novaegliae*), max POD ~ 0.54. Regions with max POD < 0.01 were masked in light grey."
#| fig-width: 7
#| fig-height: 15

depth_max_detect_clean / ci95_plot_clean

```

In addition to examining geographic patterns in the depth of maximum POD, we examined uncertainty around these values by mapping the range of depths that fell within the 50% confidence interval of maximum POD (@fig-H3PODxy, bottom). In the northern portion of the study area, where Pacific white-sided dolphins are most likely to be observed near the surface, the range of depths within the 50% CI of maximum POD is neglgible (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% pull(depthWidth), na.rm = TRUE)` m). That range increases toward the south and east (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Lagenorhynchus obliquidens") %>% filter(depth < 400) %>% pull(depthWidth), na.rm = TRUE)` m), with the greatest variability in the south and east of the study area, near the region where POD is close to 0. Baird's beaked whales exhibited minimal variability in 50% CI width of maximum depth (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m), except for a small portion of the northwestern edge of the study area (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Berardius bairdii") %>% pull(depthWidth), na.rm = TRUE)` m) near the edge of the prediction grid.

The bimodal peak of maximum POD observed in humpback whale eDNA was mirrored in the range of depths within the 50% CI of maximum POD: the depth range was largest in the transition zones between the two peaks (maximum depth range = `r max(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m), and small outside of the transition zones (minimum depth range = `r min(maxPOD_depth_clipped_clean %>% filter(BestTaxon == "Megaptera novaeangliae") %>% pull(depthWidth), na.rm = TRUE)` m).
